{\bf Intrusion Detection through Provenance-Based Histograms}

### 2013-11-22 writeup/paper.tex
More to paper

{\bf TODO: TALK ABOUT OUR TECHNIQUES.} We build on TODO's work and extend ... blah blah blah ... combine techniques together ... blah blah. The contributions of this paper are:
\begin{enumerate}
\item TODO: the techniques we use to analyze provenance data
\item TODO: types of intrusions that this system can detect, and the accuracy in detecting intrusions
\item TODO: future work for using provenance data to detect intrusions
\end{enumerate}

@@

\section{Design and Implementation}
TODO: ideas:
\begin{enumerate}
\item somayaji: sequence of system calls
\item semantic file attributes: determine types of files. use simple statistics on files.
\item histograms: use simple statistics and build histogram of file attributes on a process name/file
\item local clustering: use centrality metrics
\item fuzzy rules
\item hash k-tuples of provenance events and compare bit vectors
\item backwards and forwards causality graphs
\item access compatibility $\frac{1}{2}(C_d + C_e)$
\item intersect good DAG and bad DAG and look at intersection of DAG
\end{enumerate}

@@

For our system administration attack, we compiled over 200 C files using the unexploited version of GCC, and exported the provenance from these runs. We then replaced GCC with our exploited version, and compiled a few files, and exported the provenance again. In this way, we again acquired provenance for “normal” usage, before acquiring provenance for our attempted exploit.

\subsection{Results}
TODO:
true positives.
false positives.
graphs?

>>>>

In particular, we generate statistical models of normal behavior, against which we can test a node of interest. These statiscal models are histograms of different graph statistics, and we use these histograms to generate kernel density estimations to evaluate a test node. The contributions of this paper are:
\begin{enumerate}
\item A histogram-based technique for using provenance data to detect intrustions
\item An analysis of the different intrusions that can be identified, and the types of provenance which best contribute to these detections
\item A cogent plan for future work for using provenance data to detect intrusions
\end{enumerate}

@@

\section{Design and Implementation}
Here we present our intrusion detection technique: a general and extendable statistical method for analyzing and comparing provenance data. At a high level, we process a provenance data graph, generating a model of "normal" behavior for each process/file in the graph. Using this model, we can determine the likelihood that a given "test" node is normal or whether it deviates from normal behavior and is thus potentially an exploited or malicious program.
\subsection{Models}
Our approach requires statistical models to evaluate the normalcy of a test node. The statistical model we chose to use was a kernel density estimation (KDE), which estimates the probability density function of a value with unknown distribution. To create the KDE, we needed to provide counts of a particular statistic. In this way, for each statistic that we chose to collect, we would be able to evaluate a test node's normalcy according to that statistic. For example, we could then ask "Does node $X_n$ have a statistically believable number of output processes compared to the other $X$ nodes?"
\subsection{Provenance Statistics}
As described above, in order to generate a KDE model, we need some statistic to measure. As we expected intrusions to differ in terms of the statistics that would stand out, we needed to generate a large number of statistics, and a model for each. With regards to provenance nodes directly, we compiled the following statistics:
\begin{enumerate}
\item The number of input files to the node.
\item The number of input processes to the node.
\item The number of input pipes to the node.
\item The number of output files to the node.
\item The number of output processes to the node.
\item The number of ouput pipes to the node.
\item file\_in\_version ??? XXX
\item file\_out\_version ??? XXX
\end{enumerate}
For ease of computation and comparison, we treated the different node types (e.g. file, process, pipe) separately so that we actually had different counts for "input files to a file node" and "input files to a process node", for instance.
\subsection{Centrality Metrics}
In addition to the data that can be collected on the nodes directly, we could also compute clustering statistics by considering each node in the context of the larger graph. Using the methods espoused by Macko et al. \cite{clustering}, we could compute different centrality and clustering metrics for a node, which can then be tabulated into a histogram like the input/output counts above. 
KENNY TODO: EXPAND THIS
\subsection{Implementation}
The implementation of the above techniques consists of multiple Python files, spanning about 850 lines of (commented) code. The raw BerkeleyDB database generated by PASS is provided as input, and textual/graphical results are outputted.
KENNY TODO: EXPAND THIS

@@

For our system administration attack, we compiled the PASS toolset using GCC, and exported the provenance from these runs. We then replaced GCC with our exploited version, and compiled a few files, and exported the provenance again. In this way, we again acquired provenance for “normal” usage, before acquiring provenance for our attempted exploit.

\subsection{Results}
\subsection{Provenance Data}
After running the normal and exploited programs described above, we visualized the provenance graphs using Orbiter \cite{orbiter}. After searching for the nodes named after our programs, we were able to find both the clean and exploited nodes. The clean nodes for "hello" simply read in one file through a pipe and outputted one file, whereas the exploited version clearly has a forked "/usr/bin/ls" process. Similarly, the mcrypt simulation (implemented in Python) has only an output file as a descendent normally, but when exploited has multiple extra processes/files as descendents (see Figure 1). Similarly, the normal GCC and exploited GCC displayed very different provenance graphs. 
\begin{figure*}
  \caption{Descendents of an unexploited mcrypt-style program. Simply an output file. Below, descendents of an exploited mcrypt-style program. In addition to an (offscreen) output file, there is also an outgoing /bin/sh process, with its own children.}
  \centering
    \includegraphics[width=\textwidth]{img/mcrypt.png}
\end{figure*}

\subsection{Statistical Data}
\subsection{KDE Predictions}
TODO:
true positives.
false positives.
graphs?
